# The Somatic Motor System

The somatic motor system (SMS or voluntary nervous system) is the part of the central and peripheral nervous system associated with the voluntary control of body movements via skeletal muscles.

The somatic nervous system consists of afferent nerves or sensory nerves, and efferent nerves or motor nerves and neurons in the brain and spinal cord. Afferent nerves are responsible for relaying sensation from the body to the central nervous system; efferent nerves are responsible for sending out commands from the CNS to the body, stimulating muscle contraction; they include all the non-sensory neurons connected with skeletal muscles and skin. The a- of afferent and the e- of efferent correspond to the prefixes ad- (to, toward) and ex- (out of).

Peripheral structures of the somatic motor system include skeletal muscles and neural connections with muscle tissues. Central structures include cerebral cortex, brainstem, spinal cord, pyramidal system including the upper motor neurons, extrapyramidal system, cerebellum, and the lower motor neurons in the brainstem and the spinal cord.

## Skeletal Muscle

Skeletal muscle is one of three major muscle types, the others being cardiac muscle and smooth muscle. It is a form of striated muscle tissue, which is under the voluntary control of the somatic nervous system. Most skeletal muscles are attached to bones by bundles of collagen fibers known as tendons.

Many muscles are named by the action the muscle performs. These include:

The flexor and extensor; abductor and adductor; levator and depressor; supinator and pronator; sphincter, tensor, and rotator muscles.

A flexor muscle decreases the anterior angle at a joint; an extensor increases the anterior angle at a joint.

An abductor moves a bone away from the midline; an adductor moves a bone closer to the midline.

A levator raises a structure; a depressor moves a structure down.

A supinator turns the palm of the hand up; a pronator turns the palm down.

A sphincter decreases the size of an opening; a tensor tenses a body part; a rotator turns a bone around its axis.

René Descartes (1596–1650) was one of the first to conceive a model of reciprocal innervation (in 1626) as the principle that provides for the control of agonist and antagonist muscles. Reciprocal innervation describes skeletal muscles as existing in antagonistic pairs, with contraction of one muscle producing forces opposite to those generated by contraction of the other. For example, in the human arm, the triceps acts to extend the lower arm outward while the biceps acts to flex the lower arm inward. To reach optimum efficiency, contraction of opposing muscles must be inhibited while muscles with the desired action are excited. This reciprocal innervation occurs so that the contraction of a muscle results in the simultaneous relaxation of its corresponding antagonist.

A common example of reciprocal innervation, is the effect of the nociceptive (or nocifensive) reflex, or defensive response to pain, otherwise commonly known as the withdrawal reflex; a type of involuntary action of the body to remove the body part from the vicinity of an offending object by contracting the appropriate muscles (usually flexor muscles), while relaxing the extensor muscles.

The concept of reciprocal innervation is also applicable to the eye (Sherrington's law), wherein increased innervation to an extraocular muscle is accompanied by a simultaneous decrease in innervation to its specific antagonist, such as the medial rectus and the lateral rectus in the case of an eye looking to one side of the midline. When looking outward or laterally, the lateral rectus of one eye must contract via increased innervation, while its antagonist, the medial rectus of the same eye - shall relax. The converse would occur in the other eye, both eyes demonstrating the law of reciprocal innervation.

## The Muscle Fiber

A skeletal muscle refers to multiple bundles (fascicles) of cells joined together called muscle fibers. The fibers and muscles are surrounded by connective tissue layers called fasciae. Muscle fibers, or muscle cells, are formed from the fusion of developmental myoblasts in a process known as myogenesis. Muscle fibers are cylindrical and have more than one nucleus. They also have multiple mitochondria to meet energy needs.

Muscle fibers are in turn composed of myofibrils. The myofibrils are composed of actin and myosin filaments, repeated in units called sarcomeres, which are the basic functional units of the muscle fiber. The sarcomere is responsible for the striated appearance of skeletal muscle and forms the basic machinery necessary for muscle contraction.

Skeletal muscle exhibits a distinctive banding pattern when viewed under the microscope due to the arrangement of cytoskeletal elements in the cytoplasm of the muscle fibers. The principal cytoplasmic proteins are myosin and actin (also known as "thick" and "thin" filaments, respectively) which are arranged in a repeating unit called a sarcomere. The interaction of myosin and actin is responsible for muscle contraction.

Every single organelle and macromolecule of a muscle fiber is arranged to ensure form meets function. The cell membrane is called the sarcolemma with the cytoplasm known as the sarcoplasm. In the sarcoplasm are the myofibrils. The myofibrils are long protein bundles about 1 micrometer in diameter each containing myofilaments. Pressed against the inside of the sarcolemma are the unusual flattened myonuclei. Between the myofibrils are the mitochondria.

While the muscle fiber does not have smooth endoplasmic cisternae, it contains a sarcoplasmic reticulum. The sarcoplasmic reticulum surrounds the myofibrils and holds a reserve of the calcium ions needed to cause a muscle contraction. Periodically, it has dilated end sacs known as terminal cisternae. These cross the muscle fiber from one side to the other. In between two terminal cisternae is a tubular infolding called a transverse tubule (T tubule). T tubules are the pathways for action potentials to signal the sarcoplasmic reticulum to release calcium, causing a muscle contraction. Together, two terminal cisternae and a transverse tubule form a triad.

In addition to the actin and myosin components that constitute the sarcomere, skeletal muscle fibers also contain two other important regulatory proteins, troponin and tropomyosin, that are necessary for muscle contraction to occur. These proteins are associated with actin and cooperate to prevent its interaction with myosin. Skeletal muscle cells are excitable and are subject to depolarization by the neurotransmitter acetylcholine, released at the neuromuscular junction by motor neurons.

Once a cell is sufficiently stimulated, the cell's sarcoplasmic reticulum releases ionic calcium (Ca^2+^), which then interacts with the regulatory protein troponin. Calcium-bound troponin undergoes a conformational change that leads to the movement of tropomyosin, subsequently exposing the myosin-binding sites on actin. This allows for myosin and actin ATP-dependent cross-bridge cycling and shortening of the muscle.

## The Motor Unit

Motor units are multiple muscle fibers bundled together. When a person wants to move their body to achieve a certain task, the brain sends an impulse signal that reaches the specific motor unit through the spinal cord. After receiving the signal from the brain, the motor unit contracts muscle fibers within the group to create movement. There is no partial firing in the motor unit, meaning, once the signal is detected, all the muscles fibers within the unit contract. However, there are different intensities. Since each motor unit contracts 100% of its fiber once stimulated, types of motor units that generate different force or speed are significant.

For low intensity tasks, smaller motor units with fewer muscle fibers are used. These smaller motor units are known as low threshold motor units. They consist of type I fibers that contract much slower and thus provide less force for daily basic movement such as typing on the keyboard. For more intense tasks, motor units containing Type II muscle fibers are used. These fast twitch motor units are known as high threshold motor units. The major difference between low threshold motor units (slow twitch motor unit) and high threshold motor units (fast twitch motor unit) is that high threshold motor units control more muscle fibers and contain larger muscle fibers, in comparison to low threshold motor unit. On the other hand, the main difference between the slow twitch muscle fiber (Type I) and fast twitch muscle fiber (Type II) has the same theory of the size deviations.

During an activity of lifting heavy objects such as working out with a dumbbell, not only does low-threshold motor units, but also the high threshold motor units are recruited to compensate forces required in addition to just holding a fork, in which the energy created by the low threshold motor units is sufficient to complete the job. When giving a job, the body first recruits the slow-twitch motor units following by recruit more a more fast-twitch motor units as forces required to complete the movements increase. Thus, when the body has to carry an extremely massive object, it would recruited all the available motor units to contract for the particular muscle that has been used.

### Muscle Contraction

When contracting, thin and thick filaments slide with respect to each other by using adenosine triphosphate. This pulls the Z discs closer together in a process called sliding filament mechanism. The contraction of all the sarcomeres results in the contraction of the whole muscle fiber. This contraction of the myocyte is triggered by the action potential over the cell membrane of the myocyte. The action potential uses transverse tubules to get from the surface to the interior of the myocyte, which is continuous within the cell membrane. Sarcoplasmic reticula are membranous bags that transverse tubules touch but remain separate from. These wrap themselves around each sarcomere and are filled with Ca^2+^.

Excitation of a myocyte causes depolarization at its synapses, the neuromuscular junctions, which triggers action potential. With a singular neuromuscular junction, each muscle fiber receives input from just one somatic efferent neuron. Action potential in a somatic efferent neuron causes the release of the neurotransmitter acetylcholine.

When the acetylcholine is released it diffuses across the synapse and binds to a receptor on the sarcolemma, a term unique to muscle cells that refers to the cell membrane. This initiates an impulse that travels across the sarcolemma.

When the action potential reaches the sarcoplasmic reticulum it triggers the release of Ca^2+^ from the Ca^2+^ channels. The Ca^2+^ flows from the sarcoplasmic reticulum into the sarcomere with both of its filaments. This causes the filaments to start sliding and the sarcomeres to become shorter. This requires a large amount of ATP, as it is used in both the attachment and release of every myosin head. Very quickly Ca^2+^ is actively transported back into the sarcoplasmic reticulum, which blocks the interaction between the thin and thick filament. This in turn causes the muscle cell to relax.

## Somatic Motor Neurons

A motor neuron (or motoneuron) is a neuron whose cell body is located in the motor cortex, brainstem or the spinal cord, and whose axon (fiber) projects to the spinal cord or outside of the spinal cord to directly or indirectly control effector organs, mainly muscles and glands. There are two types of motor neuron – upper motor neurons and lower motor neurons. Axons from upper motor neurons synapse onto interneurons in the spinal cord and occasionally directly onto lower motor neurons. The axons from the lower motor neurons are efferent nerve fibers that carry signals from the spinal cord to the effectors. Types of lower motor neurons are alpha motor neurons, beta motor neurons, and gamma motor neurons.

A single motor neuron may innervate many muscle fibres and a muscle fibre can undergo many action potentials in the time taken for a single muscle twitch. Innervation takes place at a neuromuscular junction and twitches can become superimposed as a result of summation or a tetanic contraction. Individual twitches can become indistinguishable, and tension rises smoothly eventually reaching a plateau

Upper motor neurons originate in the motor cortex located in the precentral gyrus. The cells that make up the primary motor cortex are Betz cells, which are a type of pyramidal cell. The axons of these cells descend from the cortex to form the corticospinal tract. 

Corticomotorneurons project from the primary cortex directly onto motor neurons in the ventral horn of the spinal cord. Their axons synapse on the spinal motor neurons of multiple muscles as well as on spinal interneurons. They are unique to primates and it has been suggested that their function is the adaptive control of the hands including the relatively independent control of individual fingers. Corticomotorneurons have so far only been found in the primary motor cortex and not in secondary motor areas.

Nerve tracts are bundles of axons as white matter, that carry action potentials to their effectors. In the spinal cord these descending tracts carry impulses from different regions. These tracts also serve as the place of origin for lower motor neurons. There are seven major descending motor tracts to be found in the spinal cord:

* Lateral corticospinal tract
* Rubrospinal tract
* Lateral reticulospinal tract
* Vestibulospinal tract
* Medial reticulospinal tract
* Tectospinal tract
* Anterior corticospinal tract

Lower motor neurons are those that originate in the spinal cord and directly or indirectly innervate effector targets. The target of these neurons varies, but in the somatic nervous system the target will be some sort of muscle fiber.

(ref:spinal) [Efferent and afferent tracts of the spinal cord](https://commons.wikimedia.org/wiki/File:Spinal_cord_tracts_-_English.svg)

```{r spinaltracts, fig.cap='(ref:spinal)', echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("./figures/cns/Spinal_cord_tracts_-_English.svg")
```

* Alpha motor neurons innervate extrafusal muscle fibers, which are the main force-generating component of a muscle. Their cell bodies are in the ventral horn of the spinal cord and they are sometimes called ventral horn cells. A single motor neuron may synapse with 150 muscle fibers on average. The motor neuron and all of the muscle fibers to which it connects is a motor unit. Motor units are split up into 3 categories: 

  * Slow (S) motor units stimulate small muscle fibers, which contract very slowly and provide small amounts of energy but are very resistant to fatigue, so they are used to sustain muscular contraction, such as keeping the body upright. They gain their energy via oxidative means and hence require oxygen. They are also called red fibers.
  * Fast fatiguing (FF) motor units stimulate larger muscle groups, which apply large amounts of force but fatigue very quickly. They are used for tasks that require large brief bursts on energy, such as jumping or running. They gain their energy via glycolytic means and hence don't require oxygen. They are called white fibers.
  * Fast fatigue-resistant motor units stimulate moderate-sized muscles groups that don't react as fast as the FF motor units, but can be sustained much longer (as implied by the name) and provide more force than S motor units. These use both oxidative and glycolytic means to gain energy.

In addition to voluntary skeletal muscle contraction, alpha motor neurons also contribute to muscle tone, the continuous force generated by noncontracting muscle to oppose stretching. When a muscle is stretched, sensory neurons within the muscle spindle detect the degree of stretch and send a signal to the CNS. The CNS activates alpha motor neurons in the spinal cord, which cause extrafusal muscle fibers to contract and thereby resist further stretching. This process is also called the stretch reflex.

* Beta motor neurons innervate intrafusal muscle fibers of muscle spindles, with collaterals to extrafusal fibres. There are two types of beta motor neurons: Slow Contracting- These innervate extrafusal fibers. Fast Contracting- These innervate intrafusal fibers.
* Gamma motor neurons innervate intrafusal muscle fibers found within the muscle spindle. They regulate the sensitivity of the spindle to muscle stretching. With activation of gamma neurons, intrafusal muscle fibers contract so that only a small stretch is required to activate spindle sensory neurons and the stretch reflex. There are two types of gamma motor neurons: Dynamic- These focus on Bag1 fibers and enhance dynamic sensitivity. Static- These focus on Bag2 fibers and enhance stretch sensitivity.
* Regulatory factors of lower motor neurons
  * Size Principle – this relates to the soma of the motor neuron. This restricts larger neurons to receive a larger excitatory signal in order to stimulate the muscle fibers it innervates. By reducing unnecessary muscle fiber recruitment, the body is able to optimize energy consumption.
  * Persistent Inward Current (PIC) – recent animal study research has shown that constant flow of ions such as calcium and sodium through channels in the soma and dendrites influence the synaptic input. An alternate way to think of this is that the post-synaptic neuron is being primed before receiving an impulse.
  * After Hyper-polarization (AHP) – A trend has been identified that shows slow motor neurons to have more intense AHPs for a longer duration. One way to remember this is that slow muscle fibers can contract for longer, so it makes sense that their corresponding motor neurons fire at a slower rate.

## Neuromuscular Junction

A single motor neuron may innervate many muscle fibres and a muscle fibre can undergo many action potentials in the time taken for a single muscle twitch. As a result, if an action potential arrives before a twitch has completed, the twitches can superimpose on one another, either through summation or a tetanic contraction. In summation, the muscle is stimulated repetitively such that additional action potentials coming from the somatic nervous system arrive before the end of the twitch. The twitches thus superimpose on one another, leading to a force greater than that of a single twitch. A tetanic contraction is caused by constant, very high frequency stimulation - the action potentials come at such a rapid rate that individual twitches are indistinguishable, and tension rises smoothly eventually reaching a plateau.

The interface between a motor neuron and muscle fiber is a specialized synapse called the neuromuscular junction. Upon adequate stimulation, the motor neuron releases a flood of acetylcholine (Ach) neurotransmitters from the axon terminals from synaptic vesicles bind with the plasma membrane. The acetylcholine molecules bind to postsynaptic receptors found within the motor end plate. Once two acetylcholine receptors have been bound, an ion channel is opened and sodium ions are allowed to flow into the cell. The influx of sodium into the cell causes depolarization and triggers a muscle action potential. T tubules of the sarcolemma are then stimulated to elicit calcium ion release from the sarcoplasmic reticulum. It is this chemical release that causes the target muscle fiber to contract.

In invertebrates, depending on the neurotransmitter released and the type of receptor it binds, the response in the muscle fiber could be either excitatory or inhibitory. For vertebrates, however, the response of a muscle fiber to a neurotransmitter can only be excitatory, in other words, contractile. Muscle relaxation and inhibition of muscle contraction in vertebrates is obtained only by inhibition of the motor neuron itself. This is how muscle relaxants work by acting on the motor neurons that innervate muscles (by decreasing their electrophysiological activity) or on cholinergic neuromuscular junctions, rather than on the muscles themselves.

## Muscle Spindles

Muscle spindles are stretch receptors within the body of a muscle that primarily detect changes in the length of the muscle. They convey length information to the central nervous system via afferent nerve fibers. This information can be processed by the brain as proprioception. The responses of muscle spindles to changes in length also play an important role in regulating the contraction of muscles, for example, by activating motor neurons via the stretch reflex to resist muscle stretch.

The muscle spindle has both sensory and motor components.

Sensory information conveyed by primary type Ia sensory fibers which spiral around muscle fibres within the spindle, and secondary type II sensory fibers
Activation of muscle fibres within the spindle by up to a dozen gamma motor neurons and to a lesser extent by one or two beta motor neurons.

Muscle spindles are found within the belly of muscles, between extrafusal muscle fibers.[b] The specialised fibers that constitute the muscle spindle are known as intrafusal fibers (as they are present within the spindle), to distinguish themselves from the fibres of the muscle itself which are called extrafusal fibers. Muscle spindles have a capsule of connective tissue, and run parallel to the extrafusal muscle fibers.[c]

Muscle spindles are composed of 5-14 muscle fibers, of which there are three types: dynamic nuclear bag fibers (bag1 fibers), static nuclear bag fibers (bag2 fibers), and nuclear chain fibers.

Primary type Ia sensory fibers (large diameter) spiral around all intrafusal muscle fibres, ending near the middle of each fibre. Secondary type II sensory fibers (medium diameter) end adjacent to the central regions of the static bag and chain fibres. These fibres send information by stretch-sensitive mechanically-gated ion-channels of the axons.

The motor part of the spindle is provided by motor neurons: up to a dozen gamma motor neurons and one or two beta motor neurons, collectively called fusimotor neurons. These activate the muscle fibres within the spindle. Gamma motor neurons supply only muscle fibres within the spindle, whereas beta motor neurons supply muscle fibres both within and outside of the spindle. Activation of the neurons causes a contraction and stiffening of the end parts of the muscle spindle muscle fibers.

Fusimotor neurons are classified as static or dynamic according to the type of muscle fibers they innervate and their effects on the responses of the Ia and II sensory neurons innervating the central, non-contractile part of the muscle spindle.

* The static axons innervate the chain or static bag2 fibers. They increase the firing rate of Ia and II afferents at a given muscle length (see schematic of fusimotor action below).
* The dynamic axons innervate the bag1 intrafusal muscle fibers. They increase the stretch-sensitivity of the Ia afferents by stiffening the bag1 intrafusal fibers.
        
Efferent nerve fibers of gamma motoneurons also terminate in muscle spindles; they make synapses at either or both of the ends of the intrafusal muscle fibers and regulate the sensitivity of the sensory afferents, which are located in the non-contractile central (equatorial) region.

### The Stretch Reflex

Wilhelm Heinrich Erb (1840–1921) and Carl Friedrich Westphal (1833–1890) simultaneously reported the patellar tendon or knee reflex in 1875 . The term knee-jerk was recorded by Sir Michael Foster in his Textbook of physiology in 1877: "Striking the tendon below the patella gives rise to a sudden extension of the leg, known as the knee-jerk."

The patellar reflex or knee-jerk (myotatic) (monosynaptic) is a clinical and classic example of the monosynaptic reflex arc. There is no interneuron in the pathway leading to contraction of the quadriceps muscle. Instead, the sensory neuron synapses directly on a motor neuron in the spinal cord. However, there is an inhibitory interneuron used to relax the antagonistic hamstring muscle (reciprocal innervation).

Striking of the patellar tendon with a reflex hammer just below the patella stretches the muscle spindle in the quadriceps muscle. This produces a signal which travels back to the spinal cord and synapses (without interneurons) at the level of L3 in the spinal cord, completely independent of higher centres. From there, an alpha motor neuron conducts an efferent impulse back to the quadriceps femoris muscle, triggering contraction. This contraction, coordinated with the relaxation of the antagonistic flexor hamstring muscle causes the leg to kick. This is a reflex of proprioception which helps maintain posture and balance, allowing to keep one's balance with little effort or conscious thought.

When a muscle is stretched, primary type Ia sensory fibers of the muscle spindle respond to both changes in muscle length and velocity and transmit this activity to the spinal cord in the form of changes in the rate of action potentials. Likewise, secondary type II sensory fibers respond to muscle length changes (but with a smaller velocity-sensitive component) and transmit this signal to the spinal cord. The Ia afferent signals are transmitted monosynaptically to many alpha motor neurons of the receptor-bearing muscle. The reflexly evoked activity in the alpha motoneurons is then transmitted via their efferent axons to the extrafusal fibers of the muscle, which generate force and thereby resist the stretch. The Ia afferent signal is also transmitted polysynaptically through interneurons (Ia inhibitory interneurons), which inhibit alpha motorneurons of antagonist muscles, causing them to relax.

The function of the gamma motor neurons is not to supplement the force of muscle contraction provided by the extrafusal fibers, but to modify the sensitivity of the muscle spindle sensory afferents to stretch. Upon release of acetylcholine by the active gamma motor neuron, the end portions of the intrafusal muscle fibers contract, thus elongating the non-contractile central portions (see "fusimotor action" schematic below). This opens stretch-sensitive ion channels of the sensory endings, leading to an influx of sodium ions. This raises the resting potential of the endings, thereby increasing the probability of action potential firing, thus increasing the stretch-sensitivity of the muscle spindle afferents.

How does the central nervous system control gamma fusimotor neurons? It has been difficult to record from gamma motoneurons during normal movement because they have very small axons. Several theories have been proposed, based on recordings from spindle afferents.

1. Alpha-gamma coactivation. Here it is posited that gamma motoneurons are activated in parallel with alpha motoneurons to maintain the firing of spindle afferents when the extrafusal muscles shorten.
2. Fusimotor set: Gamma motoneurons are activated according to the novelty or difficulty of a task. Whereas static gamma motoneurons are continuously active during routine movements such as locomotion, dynamic gamma motoneoruns tend to be activated more during difficult tasks, increasing Ia stretch-sensitivity.
3. Fusimotor template of intended movement. Static gamma activity is a "temporal template" of the expected shortening and lengthening of the receptor-bearing muscle. Dynamic gamma activity turns on and off abruptly, sensitizing spindle afferents to the onset of muscle lengthening and departures from the intended movement trajectory.

## The Motor Cortex

The motor cortex is the region of the cerebral cortex involved in the planning, control, and execution of voluntary movements. Classically the motor cortex is an area of the frontal lobe located in the posterior precentral gyrus immediately anterior to the central sulcus. 

The motor cortex can be divided into three areas:

1. The primary motor cortex is the main contributor to generating neural impulses that pass down to the spinal cord and control the execution of movement. However, some of the other motor areas in the brain also play a role in this function. It is located on the anterior paracentral lobule on the medial surface.
1. The premotor cortex is responsible for some aspects of motor control, possibly including the preparation for movement, the sensory guidance of movement, the spatial guidance of reaching, or the direct control of some movements with an emphasis on control of proximal and trunk muscles of the body. Located anterior to the primary motor cortex.
1. The supplementary motor area (or SMA), has many proposed functions including the internally generated planning of movement, the planning of sequences of movement, and the coordination of the two sides of the body such as in bi-manual coordination. Located on the midline surface of the hemisphere anterior to the primary motor cortex.
  * The posterior parietal cortex is sometimes also considered to be part of the group of motor cortical areas; however it is best to regard it as an association cortex rather than motor. It is thought to be responsible for transforming multisensory information into motor commands, and to be responsible for some aspects of motor planning, in addition to many other functions that may not be motor related.
  * The primary somatosensory cortex, especially the part called area 3a, which lies directly against the motor cortex, is sometimes considered to be functionally part of the motor control circuitry.

Other brain regions outside the cerebral cortex are also of great importance to motor function, most notably the cerebellum, the basal ganglia, pedunculopontine nucleus and the red nucleus, as well as other subcortical motor nuclei.

In the earliest work on the motor cortex, researchers recognized only one cortical field involved in motor control. Alfred Walter Campbell was the first to suggest that there might be two fields, a "primary" motor cortex and an "intermediate precentral" motor cortex. His reasons were largely based on cytoarchitectonics, or the study of the appearance of the cortex under a microscope. The primary motor cortex contains cells with giant cell bodies known as "Betz cells". These cells were mistakenly thought to be the main outputs from the cortex, sending fibers to the spinal cord. It has since been found that Betz cells account for about 2-3% of the projections from the cortex to the spinal cord, or about 10% of the projections from the primary motor cortex to the spinal cord. The specific function of the Betz cells that distinguishes them from other output cells of the motor cortex remains unknown, but they continue to be used as a marker for the primary motor cortex.

Other researchers, such as Vogt and Vogt and Otfrid Foerster also suggested that motor cortex was divided into a primary motor cortex (area 4, according to Brodmann's naming scheme) and a higher-order motor cortex (area 6 according to Korbinian Brodmann).

Wilder Penfield notably disagreed and suggested that there was no functional distinction between area 4 and area 6. In his view both were part of the same map, though area 6 tended to emphasize the muscles of the back and neck. Woolsey who studied the motor map in monkeys also believed there was no distinction between primary motor and premotor. M1 was the name for the proposed single map that encompassed both the primary motor cortex and the premotor cortex. Although sometimes "M1" and "primary motor cortex" are used interchangeably, strictly speaking, they derive from different conceptions of motor cortex organization.

Despite the views of Penfield and Woolsey, a consensus emerged that area 4 and area 6 had sufficiently different functions that they could be considered different cortical fields. Fulton helped to solidify this distinction between a primary motor cortex in area 4 and a premotor cortex in area 6. As Fulton pointed out, and as all subsequent research has confirmed, both primary motor and premotor cortex project directly to the spinal cord and are capable of some direct control of movement. Fulton showed that when the primary motor cortex is damaged in an experimental animal, movement soon recovers; when the premotor cortex is damaged, movement soon recovers; when both are damaged, movement is lost and the animal cannot recover.

The premotor cortex is now generally divided into four sections. First it is divided into an upper (or dorsal) premotor cortex and a lower (or ventral) premotor cortex. Each of these is further divided into a region more toward the front of the brain (rostral premotor cortex) and a region more toward the back (caudal premotor cortex). A set of acronyms are commonly used: PMDr (premotor dorsal, rostral), PMDc, PMVr, PMVc. Some researchers use a different terminology. Field 7 or F7 denotes PMDr; F2 = PMDc; F5=PMVr; F4=PMVc.

PMDc is often studied with respect to its role in guiding reaching. Neurons in PMDc are active during reaching. When monkeys are trained to reach from a central location to a set of target locations, neurons in PMDc are active during the preparation for the reach and also during the reach itself. They are broadly tuned, responding best to one direction of reach and less well to different directions. Electrical stimulation of the PMDc on a behavioral time scale was reported to evoke a complex movement of the shoulder, arm, and hand that resembles reaching with the hand opened in preparation to grasp.

PMDr may participate in learning to associate arbitrary sensory stimuli with specific movements or learning arbitrary response rules. In this sense it may resemble the prefrontal cortex more than other motor cortex fields. It may also have some relation to eye movement. Electrical stimulation in the PMDr can evoke eye movements and neuronal activity in the PMDr can be modulated by eye movement.

PMVc or F4 is often studied with respect to its role in the sensory guidance of movement. Neurons here are responsive to tactile stimuli, visual stimuli, and auditory stimuli. These neurons are especially sensitive to objects in the space immediately surrounding the body, in so-called peripersonal space. Electrical stimulation of these neurons causes an apparent defensive movement as if protecting the body surface. This premotor region may be part of a larger circuit for maintaining a margin of safety around the body and guiding movement with respect to nearby objects.

PMVr or F5 is often studied with respect to its role in shaping the hand during grasping and in interactions between the hand and the mouth. Electrical stimulation of at least some parts of F5, when the stimulation is applied on a behavioral time scale, evokes a complex movement in which the hand moves to the mouth, closes in a grip, orients such that the grip faces the mouth, the neck turns to align the mouth to the hand, and the mouth opens.

Mirror neurons were first discovered in area F5 in the monkey brain by Rizzolatti and colleagues. These neurons are active when the monkey grasps an object. Yet the same neurons become active when the monkey watches an experimenter grasp an object in the same way. The neurons are therefore both sensory and motor. Mirror neurons are proposed to be a basis for understanding the actions of others by internally imitating the actions using one's own motor control circuits.

### The Supplementary Motor Cortex

Penfield described a cortical motor area, the supplementary motor area (SMA), on the top or dorsal part of the cortex. Each neuron in the SMA may influence many muscles, many body parts, and both sides of the body. The map of the body in SMA is therefore extensively overlapping. SMA projects directly to the spinal cord and may play some direct role in the control of movement.

Based on early work using brain imaging techniques in the human brain, Roland suggested that the SMA was especially active during the internally generated plan to make a sequence of movements. In the monkey brain, neurons in the SMA are active in association with specific learned sequences of movement.

Others have suggested that, because the SMA appears to control movement bilaterally, it may play a role in inter-manual coordination.

Yet others have suggested that, because of the direct projection of SMA to the spinal cord and because of its activity during simple movements, it may play a direct role in motor control rather than solely a high level role in planning sequences.

On the basis of the movements evoked during electrical stimulation, it has been suggested that the SMA may have evolved in primates as a specialist in the part of the motor repertoire involving climbing and other complex locomotion.

Based on the pattern of projections to the spinal cord, it has been suggested that another set of motor areas may lie next to the supplementary motor area, on the medial (or midline) wall of the hemisphere. These medial areas are termed the cingulate motor areas. Their functions are not yet understood.

In 1870 Eduard Hitzig and Gustav Fritsch demonstrated that electrical stimulation of certain parts of the dog brain resulted in muscular contraction on the opposite side of the body.

A little later, in 1874, David Ferrier, working in the laboratory of the West Riding Lunatic Asylum at Wakefield (at the invitation of its director, James Crichton-Browne), mapped the motor cortex in the monkey brain using electrical stimulation. He found that the motor cortex contained a rough map of the body with the feet at the top (or dorsal part) of the brain and the face at the bottom (or ventral part) of the brain. He also found that when electrical stimulation was maintained for a longer time, such as for a second, instead of being discharged over a fraction of a second, then some coordinated, seemingly meaningful movements could be caused, instead of only muscle twitches.

After Ferrier's discovery, many neuroscientists used electrical stimulation to study the map of the motor cortex in many animals including monkeys, apes, and humans.

One of the first detailed maps of the human motor cortex was described in 1905 by Campbell. He did autopsies on the brains of amputees. A person who had lost an arm would over time apparently lose some of the neuronal mass in the part of the motor cortex that normally controls the arm. Likewise, a person who had lost a leg would show degeneration in the leg part of motor cortex. In this way the motor map could be established. In the period between 1919 and 1936 others mapped the motor cortex in detail using electrical stimulation, including the husband and wife team Vogt and Vogt, and the neurosurgeon Foerster.

Perhaps the best-known experiments on the human motor map were published by Penfield in 1937. Using a procedure that was common in the 1930s, he examined epileptic patients who were undergoing brain surgery. These patients were given a local anesthetic, their skulls were opened, and their brains exposed. Then, electrical stimulation was applied to the surface of the brain to map out the speech areas. In this way, the surgeon would be able to avoid any damage to speech circuitry. The brain focus of the epilepsy could then be surgically removed. During this procedure, Penfield mapped the effect of electrical stimulation in all parts of the cerebral cortex, including motor cortex.

Penfield is sometimes mistakenly considered to be the discoverer of the map in motor cortex. It was discovered approximately 70 years before his work. However, Penfield drew a picture of a human-like figure stretched over the cortical surface and used the term "homunculus" (diminutive of "homo", Latin for "man") to refer to it. It is perhaps for this reason that his work has become so popular in neuroscience.

### The Motor Cortex Map

A simple view, that is almost certainly too limited and that dates back to the earliest work on the motor cortex, is that neurons in motor cortex control movement by a feed-forward direct pathway. In that view, a neuron in motor cortex sends an axon or projection to the spinal cord and forms a synapse on a motor neuron. The motor neuron sends an electrical impulse to a muscle. When the neuron in cortex becomes active, it causes a muscle contraction. The greater the activity in motor cortex, the stronger the muscle force. Each point in motor cortex controls a muscle or a small group of related muscles. This description is only partly correct.

Most neurons in the motor cortex that project to the spinal cord synapse on interneuron circuitry in the spinal cord, not directly onto motor neurons. One suggestion is that the direct, cortico-motoneuronal projections are a specialization that allows for the fine control of the fingers.

The view that each point in motor cortex controls a muscle or a limited set of related muscles was debated over the entire history of research on the motor cortex, and was suggested in its strongest and most extreme form by Asanuma on the basis of experiments in cats and monkeys using electrical stimulation. However, almost every other experiment to examine the map, including the classic work of Ferrier and of Penfield showed that each point in motor cortex influences a range of muscles and joints. The map is greatly overlapping. The overlap in the map is generally greater in the premotor cortex and supplementary motor cortex, but even the map in the primary motor cortex controls muscles in an extensively overlapped manner. Many studies have demonstrated the overlapping representation of muscles in the motor cortex.

It is believed that as an animal learns a complex movement repertoire, the motor cortex gradually comes to coordinate among muscles.

The clearest example of the coordination of muscles into complex movement in the motor cortex comes from the work of Graziano and colleagues on the monkey brain. They used electrical stimulation on a behavioral time scale, such as for half a second instead of the more typical hundredth of a second. They found that this type of stimulation of the monkey motor cortex often evoked complex, meaningful actions. For example, stimulation of one site in cortex would cause the hand to close, move to the mouth, and the mouth to open. Stimulation of another site would cause the hand to open, rotate until the grip faced outward, and the arm to project out as if the animal were reaching. Different complex movements were evoked from different sites and these movements were mapped in the same orderly manner in all monkeys tested. Computational models showed that the normal movement repertoire of a monkey, if arranged on a sheet such that similar movements are placed near each other, will result in a map that matches the actual map found in the monkey motor cortex. This work suggests that the motor cortex does not truly contain a homunculus-type map of the body. Instead, the deeper principle may be a rendering of the movement repertoire onto the cortical surface. To the extent that the movement repertoire breaks down partly into the actions of separate body parts, the map contains a rough and overlapping body arrangement noted by researchers over the past century.

## The Cortico-Basal Ganglia-Thalamo-Cortical Loop

The cortico-basal ganglia-thalamo-cortical loop (CBGTC loop) is a system of neural circuits in the brain. The loop involves connections between the cortex, the basal ganglia, the thalamus, and back to the cortex. It is of particular relevance to hyperkinetic and hypokinetic movement disorders, such as Parkinson's disease and Huntington's disease, as well as to mental disorders of control, such as attention deficit hyperactivity disorder (ADHD), obsessive–compulsive disorder (OCD), and Tourette syndrome.

The CBGTC loop primarily consists of modulatory dopaminergic projections from the pars compacta of the substantia nigra, and ventral tegmental area as well as excitatory glutamatergic projections from the cortex to the striatum, where these projections form synapses with excitatory and inhibitory pathways that relay back to the cortex. The loop was originally proposed as a part of a model of the basal ganglia called the parallel processing model, which has been criticized and modified into another model called the center surround model.

Current organization schemes characterize cortico-basal ganglia interactions as segregated parallel processing, meaning there is little convergence of distinct cortical areas in the basal ganglia. This is thought to explain the topographically organized functionality of the striatum. The striatum is organized on a rostro-caudal axis, with the rostral putamen and caudate serving associative and cognitive functions and the caudal areas serving sensorimotor function. Sometimes when the striatum is the expressed target the loop is referred to as the cortico-striatal-thalamic-cortical loop.

The two major input structures of the circuit are the striatum and the subthalamic nucleus (STN). The striatum receives inputs from both the cortex and the pars compacta of the substantia nigra (SNc), while the STN only receives cortical inputs.

Two pathways emerge from the striatum. One pathway is called the indirect (or NoGo) pathway and is inhibitory. This projects to and inhibits the globus pallidus externus (GPe), resulting in the disinhibition of the globus pallidus internus (GPi), leading to inhibition of the thalamus. This pathway also, as a result of inhibiting the GPe, disinhibits the subthalamic nucleus, which results in excitation of the GPi, and therefore inhibition of the thalamus.

The second pathway, is called the direct (or Go) pathway and is excitatory. This pathway inhibits the GPi, resulting in the disinhibition of the thalamus. The direct pathway mostly consists of monosynaptic connections driven by dopamine receptor D1, adenosine A1 receptor, and muscarinic acetylcholine receptor M4, while the indirect pathway relies on connections driven by dopamine receptor D2, adenosine A2A receptor, and muscarinic acetylcholine receptor M1.

(ref:circuits) [Circuits of the basal ganglia.](https://commons.wikimedia.org/wiki/File:Basal_ganglia_circuits.svg) Picture shows 2 coronal slices that have been superimposed to include the involved basal ganglia structures. + and - signs at the point of the arrows indicate respectively whether the pathway is excitatory or inhibitory in effect. Green arrows refer to excitatory glutamatergic pathways, red arrows refer to inhibitory GABAergic pathways and turquoise arrows refer to dopaminergic pathways that are excitatory on the direct pathway and inhibitory on the indirect pathway. Note that dis-inhibitory pathways in effect are excitatory on the feedback to the cortex, while dis-dis-inhibitory pathways in effect are inhibitory. See en:Basal ganglia#Connections

```{r basalcircuits, fig.cap='(ref:circuits)', echo=FALSE, message=FALSE, warning=FALSE}
knitr::include_graphics("./figures/motor/Basal_ganglia_circuits.svg")
```

The parallel CBGTC loops have been segregated according to the functions of associated cortical regions. One scheme involves the division into limbic and motor loops, with the motor loops containing indirect and direct pathways, which are in turn interconnected with the limbic loop that projects into the ventral striatum. The loop has also been divided into limbic, associative, oculomotor, and motor circuits to explain the role of dopamine in the basal ganglia on motivational states. A five loop division based on primary cortical targets has been described as follows:

* A motor circuit originating in the supplementary motor area, motor cortex, and somatosensory cortex, which in turn projects to the putamen, which projects to the ventrolateral GPi and caudolateral SNr, before returning to the cortex via the ventralis lateralis pars oralis and ventralis lateralis pars medialis.
* An oculomotor circuit originating in the frontal eye fields projecting to the body of the caudate, and returning via the caudal dorsomedial GPi/ventromedial SNr, and then the lateral ventralis anterior pars magnocellularis and medialis dorsalis pars paralarnellaris.
* A dorsolateral prefrontal circuit involving projections from the dlPFC and posterior parietal cortex, that projects to the dorsolateral head of the caudate, which in turn projects to the lateral dorsomedial GPi/rostrolateral SNr, which projects to the ventralis anterior pars parvocellularis and medialis dorsalis pars parvocellularis.
* A lateral orbitofrontal circuit projecting to the ventromedial caudate head, which projects through the medial dorsomedial GPi/rostromedial SNr to the medial ventralis medialis pars magnocellularis/medialis dorsalis pars magnocellularis.
* An anterior cingulate circuit that involves projections from the ACC to the ventral striatum, through the rostolateral GPi/VP/rostrodorsal SNr, which returns via the posteromedial medialis dorsalis.

A problem identified with the current anatomy of the circuit is that the time delay between the direct and indirect pathways should result in this circuit not working. To overcome this, the center surround hypothesis posits a hyperdirect pathway from the cortex would inhibit other inputs besides one focused cortical input. However, the timing of basal ganglia activity and limb moment, as well as lesion studies do not support this hypothesis

Two models have been proposed to explain how actions are selected in the basal ganglia. The actor-critic model suggests that actions are generated and evaluated by a "critic" in the ventral striatum, while the actions are carried out by an "actor" in the dorsal striatum. Another model proposes the basal ganglia acts as a selection mechanism, where actions are generated in the cortex and are selected based on context by the basal ganglia. The CBGTC loop is also involved in reward discounting, with firing increasing with an unexpected or greater than expected reward. One review supported the idea that the cortex was involved in learning actions regardless of their outcome, while the basal ganglia was involved in selecting appropriate actions based on associative reward based trial and error learning.

The CBGTC loop has been implicated in many diseases. For example, in Parkinson's disease, degeneration of dopaminergic neurons leading to decreased activity of the excitatory pathway is thought to result in hypokinesia, and in Huntington's disease, degeneration of GABAergic neurons driving the inhibitory pathway is thought to result in the jerky body movements. The co-degeneration of limbic projections along with motor projections may result in many of the psychiatric symptoms of these primarily motor illnesses. 

## The Pyramidal Motor System

The pyramidal motor system, also called the pyramidal tract or the corticospinal tract, start in the motor center of the cerebral cortex. There are upper and lower motor neurons in the corticospinal tract. The motor impulses originate in the giant pyramidal cells or Betz cells of the motor area; i.e., precentral gyrus of cerebral cortex. These are the upper motor neurons (UMN) of the corticospinal tract. The axons of these cells pass in the depth of the cerebral cortex to the corona radiata and then to the internal capsule passing through the posterior branch of internal capsule and continue to descend in the midbrain and the medulla oblongata. In the lower part of Medulla oblongata 80 to 85% of these fibers decussate (pass to the opposite side) and descend in the white matter of the lateral funiculus of the spinal cord on the opposite side. The remaining 15 to 20% pass to the same side. Fibers for the extremities (limbs) pass 100% to the opposite side. The fibers of the corticospinal tract terminate at different levels in the anterior horn of the grey matter of the spinal cord. Here the lower motor neurons (LMN) of the corticospinal cord are located. Peripheral motor nerves carry the motor impulses from the anterior horn to the voluntary muscles.

The pyramidal tracts include both the corticobulbar tract and the corticospinal tract. These are aggregations of efferent nerve fibers from the upper motor neurons that travel from the cerebral cortex and terminate either in the brainstem (corticobulbar) or spinal cord (corticospinal) and are involved in the control of motor functions of the body.

The corticobulbar tract conducts impulses from the brain to the cranial nerves. These nerves control the muscles of the face and neck and are involved in facial expression, mastication, swallowing, and other functions.

The corticospinal tract conducts impulses from the brain to the spinal cord. It is made up of a lateral and anterior tract. The corticospinal tract is involved in voluntary movement. The majority of fibres of the corticospinal tract cross over in the medulla oblongata, resulting in muscles being controlled by the opposite side of the brain. The corticospinal tract also contains the axons of Betz cells (the largest pyramidal cells) located in the primary motor cortex.

The pyramidal tracts are named because they pass through the pyramids of the medulla oblongata. The corticospinal fibers when descending from the internal capsule to the brain stem, converge to a point from multiple directions giving the impression of an inverted pyramid.

The myelination of the pyramidal fibres is incomplete at birth and gradually progresses in cranio-caudal direction and thereby progressively gaining functionality. Most of the myelination is complete by two years of age and thereafter it progresses very slowly in cranio-caudal direction up to twelve years of age.

The term pyramidal tracts refers to upper motor neurons that originate in the cerebral cortex and terminate in the spinal cord (corticospinal) or brainstem (corticobulbar). Nerves emerge in the cerebral cortex, pass down and may cross sides in the medulla oblongata, and travel as part of the spinal cord until they synapse with interneurons in the grey column of the spinal cord.

There is some variation in terminology. The pyramidal tracts definitively encompass the corticospinal tracts, and many authors also include the corticobulbar tracts.

### The Corticospinal Tract

Nerve fibres in the corticospinal tract originate from pyramidal cells in layer V of the cerebral cortex. Fibres arise from the primary motor cortex (about 30%), supplementary motor area and the premotor cortex (together also about 30%), and the somatosensory cortex, parietal lobe, and cingulate gyrus supplies the rest. The cells have their bodies in the cerebral cortex, and the axons form the bulk of the pyramidal tracts. The nerve axons travel from the cortex through the posterior limb of internal capsule, through the cerebral peduncle and into the brainstem and anterior medulla oblongata. Here they form two prominences called the medulla oblongatary pyramids. Below the prominences, the majority of axons cross over to the opposite side from which they originated, known as decussation. The axons that cross over move to the outer part of the medulla oblongata and form the lateral corticospinal tract, whereas the fibres that remain form the anterior corticospinal tract. About 80% of axons cross over and form the lateral corticospinal tract; 10% do not cross over and join the tract, and 10% of fibres travel in the anterior corticospinal tract.

The nerve axons traveling down the tract are the efferent nerve fibers of the upper motor neurons. These axons travel down the tracts in the white matter of the spinal cord until they reach the vertebral level of the muscle that they will innervate. At this point, the axons synapse with lower motor neurons. The majority of axons do not directly synapse with lower motor neurons, but instead synapse with an interneuron that then synapses with a lower motor neuron. This generally occurs in the anterior grey column. Nerve axons of the lateral corticospinal tract that did not cross over in the medulla oblongata do so at the level of the spinal cord they terminate in.

These tracts contain more than 1 million axons and the majority of the axons are myelinated. The corticospinal tracts myelinate largely during the first and second years after birth. The majority of nerve axons are small (<4μm) in diameter. About 3% of nerve axons have a much larger diameter (16μm) and arise from Betz cells, mostly in the leg area of the primary motor cortex. These cells are notable because of their rapid conduction rate, over 70m/sec, the fastest conduction of any signals from the brain to the spinal cord.

### The Corticobulbar Tract

Fibres from the ventral motor cortex travel with the corticospinal tract through the internal capsule, but terminate in a number of locations in the midbrain (cortico-mesencephalic tract), pons (Corticopontine tract), and medulla oblongata (cortico-bulbar tract). The upper motor neurons of the corticobulbar tract synapse with interneurons or directly with the lower motor neurons located in the motor cranial nerve nuclei, namely oculomotor, trochlear, motor nucleus of the trigeminal nerve, abducens, facial nerve and accessory and in the nucleus ambiguus to the hypoglossal, vagus and accessory nerves. These nuclei are supplied by nerves from both sides of the brain, with the exception of the parts of the facial nerve that control muscles of the lower face. These muscles are only innervated by nerves from the contralateral (opposite) side of the cortex.

The nerves within the corticospinal tract are involved in movement of muscles of the body. Because of the crossing-over of fibres, muscles are supplied by the side of the brain opposite to that of the muscle. The nerves within the corticobulbar tract are involved in movement in muscles of the head. They are involved in swallowing, phonation, and movements of the tongue. By virtue of involvement with the facial nerve, the corticobulbar tract is also responsible for transmitting facial expression. With the exception of lower muscles of facial expression, all functions of the corticobulbar tract involve inputs from both sides of the brain.

## The Extrapyramidal Motor System

The extrapyramidal motor system consists of motor-modulation systems, particularly the basal ganglia and cerebellum. The system is called extrapyramidal to distinguish it from the tracts of the motor cortex that reach their targets by traveling through the pyramids of the medulla. The pyramidal tracts (corticospinal tract and corticobulbar tracts) may directly innervate motor neurons of the spinal cord or brainstem (anterior (ventral) horn cells or certain cranial nerve nuclei), whereas the extrapyramidal system centers on the modulation and regulation (indirect control) of anterior (ventral) horn cells.

Extrapyramidal tracts are chiefly found in the reticular formation of the pons and medulla, and target lower motor neurons in the spinal cord that are involved in reflexes, locomotion, complex movements, and postural control. These tracts are in turn modulated by various parts of the central nervous system, including the nigrostriatal pathway, the basal ganglia, the cerebellum, the vestibular nuclei, and different sensory areas of the cerebral cortex. All of these regulatory components can be considered part of the extrapyramidal system, in that they modulate motor activity without directly innervating motor neurons.

The extrapyramidal tracts include parts of the following:

* rubrospinal tract
* pontine reticulospinal tract
* medullary reticulospinal tract
* lateral vestibulospinal tract
* tectospinal tract

## The Basal Ganglia

The basal ganglia (or basal nuclei) are a group of subcortical nuclei, of varied origin, in the brains of vertebrates, including humans, which are situated at the base of the forebrain and top of the midbrain. There are some differences in the basal ganglia of primates. Basal ganglia are strongly interconnected with the cerebral cortex, thalamus, and brainstem, as well as several other brain areas. The basal ganglia are associated with a variety of functions, including control of voluntary motor movements, procedural learning, habit learning, eye movements, cognition, and emotion.

The nomenclature of the basal ganglia system and its components has always been problematic. Early anatomists, seeing the macroscopic anatomical structure but knowing nothing of the cellular architecture or neurochemistry, grouped together components that are now believed to have distinct functions (such as the internal and external segments of the globus pallidus), and gave distinct names to components that are now thought to be functionally parts of a single structure (such as the caudate nucleus and putamen).

The term "basal" comes from the fact that most of its elements are located in the basal part of the forebrain. The term ganglia is a misnomer: In modern usage, neural clusters are called "ganglia" only in the peripheral nervous system; in the central nervous system they are called "nuclei". For this reason, the basal ganglia are also occasionally known as the "basal nuclei". 

The main components of the basal ganglia – as defined functionally – are the striatum; both dorsal striatum (caudate nucleus and putamen) and ventral striatum (nucleus accumbens and olfactory tubercle), globus pallidus, ventral pallidum, substantia nigra, and subthalamic nucleus. Each of these components has a complex internal anatomical and neurochemical organization. The largest component, the striatum (dorsal and ventral), receives input from many brain areas beyond the basal ganglia, but only sends output to other components of the basal ganglia. The pallidum receives input from the striatum, and sends inhibitory output to a number of motor-related areas. The substantia nigra is the source of the striatal input of the neurotransmitter dopamine, which plays an important role in basal ganglia function. The subthalamic nucleus receives input mainly from the striatum and cerebral cortex, and projects to the globus pallidus.

Popular theories implicate the basal ganglia primarily in action selection – in helping to decide which of several possible behaviors to execute at any given time. In more specific terms, the basal ganglia's primary function is likely to control and regulate activities of the motor and premotor cortical areas so that voluntary movements can be performed smoothly. Experimental studies show that the basal ganglia exert an inhibitory influence on a number of motor systems, and that a release of this inhibition permits a motor system to become active. The "behavior switching" that takes place within the basal ganglia is influenced by signals from many parts of the brain, including the prefrontal cortex, which plays a key role in executive functions.

The basal ganglia are of major importance for normal brain function and behaviour. Their dysfunction results in a wide range of neurological conditions including disorders of behaviour control and movement. Those of behaviour include Tourette syndrome, obsessive–compulsive disorder, and addiction. Movement disorders include, most notably Parkinson's disease, which involves degeneration of the dopamine-producing cells in the substantia nigra, Huntington's disease, which primarily involves damage to the striatum, dystonia, and more rarely hemiballismus. The basal ganglia have a limbic sector whose components are assigned distinct names: the nucleus accumbens, ventral pallidum, and ventral tegmental area (VTA). There is considerable evidence that this limbic part plays a central role in reward learning as well as cognition and frontal lobe functioning, via the mesolimbic pathway from the VTA to the nucleus accumbens that uses the neurotransmitter dopamine, and the mesocortical pathway. A number of highly addictive drugs, including cocaine, amphetamine, specific medications that are prescribed by a doctor, and nicotine, are thought to work by increasing the efficacy of this dopamine signal. There is also evidence implicating overactivity of the VTA dopaminergic projection in schizophrenia.

The basal ganglia form a fundamental component of the cerebrum. In contrast to the cortical layer that lines the surface of the forebrain, the basal ganglia are a collection of distinct masses of gray matter lying deep in the brain not far from the junction of the thalamus. They lie to the side of and surround the thalamus. Like most parts of the brain, the basal ganglia consist of left and right sides that are virtual mirror images of each other.

In terms of anatomy, the basal ganglia are divided into four distinct structures, depending on how superior or rostral they are (in other words depending on how close to the top of the head they are): Two of them, the striatum and the pallidum, are relatively large; the other two, the substantia nigra and the subthalamic nucleus, are smaller. In the illustration to the right, two coronal sections of the human brain show the location of the basal ganglia components. Of note, and not seen in this section, the subthalamic nucleus and substantia nigra lie farther back (posteriorly) in the brain than the striatum and pallidum.

### The Striatum

The striatum is a subcortical structure generally divided into the dorsal striatum and ventral striatum, although a medial lateral classification has been suggested to be more relevant behaviorally and is being more widely used.

The striatum is composed mostly of medium spiny neurons. These GABAergic neurons project to the external (lateral) globus pallidus and internal (medial) globus pallidus as well as the substantia nigra pars reticulata. The projections into the globus pallidus and substantia nigra are primarily dopaminergic, although enkephalin, dynorphin and substance P are expressed. The striatum also contains interneurons that are classified into nitrergic neurons (due to use of nitric oxide as a neurotransmitter), tonically active[clarification needed] cholinergic interneurons, parvalbumin-expressing neurons and calretinin-expressing neurons. The dorsal striatum receives significant glutamatergic inputs from the cortex, as well as dopaminergic inputs from the substantia nigra pars compacta. The dorsal striatum is generally considered to be involved in sensorimotor activities. The ventral striatum receives glutamatergic inputs from the limbic areas as well as dopaminergic inputs from the VTA, via the mesolimbic pathway. The ventral striatum is believed to play a role in reward and other limbic functions. The dorsal striatum is divided into the caudate and putamen by the internal capsule while the ventral striatum is composed of the nucleus accumbens and olfactory tubercle. The caudate has three primary regions of connectivity, with the head of the caudate demonstrating connectivity to the prefrontal cortex, cingulate cortex and amygdala. The body and tail show differentiation between the dorsolateral rim and ventral caudate, projecting to the sensorimotor and limbic regions of the striatum respectively. Striatopallidal fibres connect the striatum to the pallidus.

### The  Pallidum

The pallidum consists of a large structure called the globus pallidus ("pale globe") together with a smaller ventral extension called the ventral pallidum. The globus pallidus appears as a single neural mass, but can be divided into two functionally distinct parts, called the internal (or medial) and external (lateral) segments, abbreviated GPi and GPe. Both segments contain primarily GABAergic neurons, which therefore have inhibitory effects on their targets. The two segments participate in distinct neural circuits. The GPe receives input mainly from the striatum, and projects to the subthalamic nucleus. The GPi receives signals from the striatum via the "direct" and "indirect" pathways. Pallidal neurons operate using a disinhibition principle. These neurons fire at steady high rates in the absence of input, and signals from the striatum cause them to pause or reduce their rate of firing. Because pallidal neurons themselves have inhibitory effects on their targets, the net effect of striatal input to the pallidum is a reduction of the tonic inhibition exerted by pallidal cells on their targets (disinhibition) with an increased rate of firing in the targets.

### The Substantia nigra

The substantia nigra is a midbrain gray matter portion of the basal ganglia that has two parts – the pars compacta (SNc) and the pars reticulata (SNr). SNr often works in unison with GPi, and the SNr-GPi complex inhibits the thalamus. Substantia nigra pars compacta (SNc) however, produces the neurotransmitter dopamine, which is very significant in maintaining balance in the striatal pathway. The circuit portion below explains the role and circuit connections of each of the components of the basal ganglia.

The substantia nigra (SN) is a basal ganglia structure located in the midbrain that plays an important role in reward and movement. Substantia nigra is Latin for "black substance", reflecting the fact that parts of the substantia nigra appear darker than neighboring areas due to high levels of neuromelanin in dopaminergic neurons. It was discovered in 1784 by Félix Vicq-d'Azyr, and Samuel Thomas von Sömmerring alluded to this structure in 1791. Parkinson's disease is characterized by the loss of dopaminergic neurons in the substantia nigra pars compacta.

Although the substantia nigra appears as a continuous band in brain sections, anatomical studies have found that it actually consists of two parts with very different connections and functions: the pars compacta (SNpc) and the pars reticulata (SNpr). This classification was first proposed by Sano in 1910. The pars compacta serves mainly as an output to the basal ganglia circuit, supplying the striatum with dopamine. The pars reticulata, though, serves mainly as an input, conveying signals from the basal ganglia to numerous other brain structures.

The substantia nigra is an important player in brain function, in particular, in eye movement, motor planning, reward-seeking, learning, and addiction. Many of the substantia nigra's effects are mediated through the striatum. The nigral dopaminergic input to the striatum via the nigrostriatal pathway is intimately linked with the striatum's function. The co-dependence between the striatum and substantia nigra can be seen in this way: when the substantia nigra is electrically stimulated, no movement occurs; however, the symptoms of nigral degeneration due to Parkinson's is a poignant example of the substantia nigra's influence on movement. In addition to striatum-mediated functions, the substantia nigra also serves as a major source of GABAergic inhibition to various brain targets.

The substantia nigra is critical in the development of many diseases and syndromes, including parkinsonism and Parkinson's disease.
Parkinson's disease is a neurodegenerative disease characterized, in part, by the death of dopaminergic neurons in the SNpc. The major symptoms of Parkinson's disease include tremor, akinesia, bradykinesia, and stiffness. Other symptoms include disturbances to posture, fatigue, sleep abnormalities, and depressed mood. The cause of death of dopaminergic neurons in the SNpc is unknown. The substantia nigra is the target of chemical therapeutics for the treatment of Parkinson's disease. Levodopa (commonly referred to as L-DOPA), the dopamine precursor, is the most commonly prescribed medication for Parkinson's disease, despite controversy concerning the neurotoxicity of dopamine and L-DOPA. The drug is especially effective in treating patients in the early stages of Parkinson's, although it does lose its efficacy over time. Levodopa can cross the blood–brain barrier and increases dopamine levels in the substantia nigra, thus alleviating the symptoms of Parkinson's disease. The drawback of levodopa treatment is that it treats the symptoms of Parkinson's (low dopamine levels), rather than the cause (the death of dopaminergic neurons in the substantia nigra).

### The Subthalamic Nucleus

The subthalamic nucleus is a diencephalic gray matter portion of the basal ganglia, and the only portion of the ganglia that produces an excitatory neurotransmitter, glutamate. The role of the subthalamic nucleus is to stimulate the SNr-GPi complex and it is part of the indirect pathway. The subthalamic nucleus receives inhibitory input from the external part of the globus pallidus and sends excitatory input to the GPi.
